---
title: "Анализ данных ген.экспрессии гранулезных клеток при СПКЯ"
author: "Игорь Буланов"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Анализ ген.экспресии

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
```

# Предобработка данных исследования GSE193123

```{r, echo = FALSE}
library(data.table)
cts <- read.csv("/home/ibulanov/data/PCOS/GEO/GSE168404_mRNA_C-vs-P.all.txt")
cts <- cts[,c(1:7)]
cts <- column_to_rownames(cts, var = "gene_id")
#colnames(cts) <- gsub("^1_", "", colnames(cts))

coldata <- data.frame(batch = colnames(cts), condition = 
                        c(rep("CONTROL", 3),
                          rep("PCOS", 3)
                          ))

dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design= ~condition)

```

# Exploratory analysis and visualization

## Pre-filtering the dataset

```{r}
nrow(dds)
keep <- rowSums(counts(dds)) > 1
dds <- dds[keep,]
nrow(dds)
```

##The variance stabilizing transformation and the rlog

```{r, echo=FALSE}
vsd <- vst(dds, blind = FALSE)
head(assay(vsd), 3)

rld <- rlog(dds, blind = FALSE)
head(assay(rld), 3)

colData(vsd)

library("dplyr")
library("ggplot2")

dds <- estimateSizeFactors(dds)

df <- bind_rows(
  as_tibble(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_tibble(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
  as_tibble(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

## Sample distances

```{r, echo=FALSE}
sampleDists <- dist(t(assay(vsd)))
sampleDists

library("pheatmap")
library("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste( vsd$dex, vsd$cell, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)

library("PoiClaClu")
poisd <- PoissonDistance(t(counts(dds)))

samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( dds$dex, dds$cell, sep=" - " )
colnames(samplePoisDistMatrix) <- NULL
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
```

#PCA plot

```{r}
plotPCA(vsd, intgroup = "condition") +
  ggtitle("PCA of GSE193123")

#PCA with VST data
pcaData <- plotPCA(vsd, intgroup = c( "batch", "condition"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = batch, shape = condition)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")

#GLM-PCA
library("glmpca")
gpca <- glmpca(counts(dds), L=2)
gpca.dat <- gpca$factors
gpca.dat$dim1 <- dds$batch
gpca.dat$dim2 <- dds$condition
ggplot(gpca.dat, aes(x = dim1, y = dim2, color = coldata$batch, shape = coldata$condition)) +
  geom_point(size =3) + coord_fixed() + ggtitle("glmpca - Generalized PCA")


#MDS Plot
mds <- as.data.frame(colData(vsd))  %>%
         cbind(cmdscale(sampleDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = batch, shape = condition)) +
  geom_point(size = 3) + coord_fixed() + ggtitle("MDS with VST data")

#MDS with PoissonDistance
mdsPois <- as.data.frame(colData(dds)) %>%
   cbind(cmdscale(samplePoisDistMatrix))
ggplot(mdsPois, aes(x = `1`, y = `2`, color = batch, shape = condition)) +
  geom_point(size = 3) + coord_fixed() + ggtitle("MDS with PoissonDistances")
```

#Анализ диф.экспрессии

```{r, eval=FALSE}
dds <- DESeq(dds)
res <- results(dds)
#res <- results(dds, contrast=c("dex","trt","untrt"))
mcols(res, use.names = TRUE)
summary(res)
```

# Конвертация имен генов и сохранение

```{r}
res_tbl <- as.data.frame(res@listData)
rownames(res_tbl) <- rownames(res)
res_tbl <- rownames_to_column(res_tbl, var = "gene")

library(biomaRt)
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

#convert to external gene name
gene_names <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_gene_id",
  values = res_tbl$gene,
  mart = ensembl
)

#connect new names to old ones
resTblMerged <- merge(gene_names, res_tbl,
                        by.x = "ensembl_gene_id",
                        by.y = "gene", all=TRUE)
#if there is NA gene symbol, leave ensembl gene ids
resTblMerged$external_gene_name <-
  ifelse(is.na(resTblMerged$external_gene_name),
             resTblMerged$ensembl_gene_id,
       resTblMerged$external_gene_name
       )
#if there is empty space isntead of gene symbol, leave ensembl gene ids
resTblMerged$external_gene_name <-
  ifelse(resTblMerged$external_gene_name== "",
             resTblMerged$ensembl_gene_id,
       resTblMerged$external_gene_name
       )
#remove first column (ensembl gene ids)
resTblMerged <- resTblMerged[,-1]


#write_delim(res_tbl, "DGE_result_GSE193123.tsv", delim = '\t')

```

#Визуализация (Вулканический плот)

```{r}
library(EnhancedVolcano)
#Plot
GSE193123_EV <- EnhancedVolcano(resTblMerged[,-1],
  lab = resTblMerged$external_gene_name,
  x = 'log2FoldChange',
  y = 'pvalue',
  title = "GSE193123")
```

#Разделение списка генов на overexpressed и underexpressed
```{r}
#separate for two tables (over and under exp genes)
diffExprGenes <- subset(resTblMerged, pvalue < 0.05 & abs(log2FoldChange) > 1)

#filter under- and overexpressed genes separately
underExprGenes <- subset(resTblMerged, pvalue < 0.05 & log2FoldChange < -1) |> arrange(log2FoldChange) |> dplyr::select(external_gene_name, log2FoldChange, pvalue)
  
overExprGenes <- subset(resTblMerged, pvalue < 0.05 & log2FoldChange > 1) |> arrange(desc(log2FoldChange)) |> dplyr::select(external_gene_name, log2FoldChange, pvalue)

#save
write_delim(underexpGenesOut, "GSE193123_UnderExpressedGenes.tsv", delim = '\t')
write_delim(overExprGenes, "GSE193123_OverExpressedGenes.tsv", delim = '\t')
```

